\chapter{Advanced Usage} \label{chap:advanced-usage}

While Chapter \ref{chap:basic-usage} outlined the basic usage of {\ViennaData}, this Chapter focuses
on customizing the library to the user's demands and the particular environment. It is explained how
the object identification can be changed, how the internal storage of data can be customized and data
accesses can be tuned for maximum speed.

\section{Retrieving IDs from Objects} \label{sec:id-retrieval}
In order to associate data with objects, one has to distinguish between different objects. The only portable identification mechanism
is the address of an object, thus this is the default identification mechanism {\ViennaData}.
However, object addresses can be randomly scattered over the address space, which leads to tree-based data structures in order to
provide a reasonably fast lookup of data.

However, it can be the case that there are additional means for identification provided.
Let us consider a class \lstinline|MyClass| equipped with a member function \lstinline|id()| that
is used to identify objects. The following code configures {\ViennaData} to use this member function
as identification mechanism for all objects
\begin{lstlisting}
namespace viennadata
{
 namespace traits
 {
  template <>
  struct id<MyClass>
  {
   typedef long   id_type;
   static id_type get(MyClass const & obj) { return obj.id(); }
  };
 }
}
\end{lstlisting}
Summing up, the library user has to provide a specialization for the template class \lstinline|id|
for each class and provide a type definition for the ID-type \lstinline|id_type| and provide a static
function that returns the ID for an object of the respective class.

\section{Dense Data} \label{sec:dense-data}
By default, {\ViennaData} internally uses a tree-based layout similar to
\begin{lstlisting}
 std::map<ObjType *, DataContainer>
\end{lstlisting}
to store and retrieve data for objects of type \lstinline|MyClass| by the use of a data container type \lstinline|DataContainer|.
This induces lookup-costs of the order $\log N$, where $N$ denotes the number of objects for which data is stored.

However, if the range of IDs of objects of a type \lstinline|MyClass| is known to be in the range $0, \ldots, N]$,
a tree-based layout can be configured to be of type
\begin{lstlisting}
 std::vector<DataContainer>
\end{lstlisting}
This requires that a custom identification mechanism is set up as outlined in the previous section. Moreover, the 
custom \lstinline|id_type| must be compatible with being used as an array index. Since such a storage scheme is only memory efficient
if the number of objects is comparable to $N$, we refer to this scheme as \emph{dense data storage}.
To change the default tree-based data storage scheme to the dense storage scheme for objects of type \lstinline|MyClass|, the lines
\begin{lstlisting}
namespace viennadata
{
 namespace traits
 {
  template <typename KeyType, typename ValueType>
  struct storage<KeyType, ValueType, MyClass>
  {
   typedef dense_data_tag    tag;
  };
 }
}
\end{lstlisting}
are sufficient. Due to the template specialization, a fine-grained control over which storage to use for the various types of data and keys is readily provided.

While tree-based schemes offer automatic memory management, this is not the case for the dense data storage. In this case, 
the user has to reserve the required memory manually \emph{prior to any other calls} of {\ViennaData} functions.
To reserve data memory for data of type \lstinline|ValueType| with keys of type \lstinline|KeyType| for 42 objects of the same type as \lstinline|obj|,
the code
\begin{lstlisting}
  viennadata::reserve<KeyType,ValueType>(42)(obj);
\end{lstlisting}
is required. Similarly, memory for other data, key and object types is reserved.

\NOTE{The dense storage scheme should only be used if either
memory consumption or data access times turn out to be critical.}

\section{Register Data} \label{sec:register-data}
The use of \lstinline|vienndata::all| for the \lstinline|copy()|, \lstinline|move()| and \lstinline|erase()| functions as described in Chapter \ref{chap:basic-usage}
requires the tracking of types used at compile time during run time. This is by default handled automatically in {\ViennaData} at the cost of a small overhead at data access.
For performance-critical applications, this automatism can be disabled for each triple of key type \lstinline|KeyType|, data type \lstinline|ValueType| and object type \lstinline|ObjectType| by
\begin{lstlisting}
namespace viennadata
{
 namespace traits
 {
  template <>
  struct signup<KeyType, ValueType, MyClass>
  {
   typedef manual_signup_tag    tag;
  };
 }
}
\end{lstlisting}
Again, the partial specialization can be adjusted to match a larger class of types.

Once the automatic tracking of types at runtime is disabled, the correct use of \lstinline|viennadata::all| requires that all each triple of key type \lstinline|KeyType|, data type \lstinline|ValueType| and object type \lstinline|ObjectType| is registered manually as
\begin{lstlisting}
  viennadata::signup<KeyType,ValueType>()(obj);
\end{lstlisting}
Here, \lstinline|obj| is of type \lstinline|ObjType|. Note that multiple calls of \lstinline|viennadata::signup()| with the same types has the same effect as a single call.

\section{Type-Based Key Dispatch} \label{sec:compiletime-keys}
As already shown in the Chapter \ref{chap:basic-usage}, {\ViennaData} can be configured to perform a purely type-based key dispatch for data access.
The impact on performance can be considerable, especially when compared to potentially costly string comparisons.
To enable a compile time dispatch for a key of type \lstinline|KeyType|, the lines
\begin{lstlisting}
namespace viennadata
{
 namespace traits
 {
  template <>
  struct dispatch<KeyType>
  {
    typedef type_key_dispatch_tag    tag;
  };
 }
} 
\end{lstlisting}
are required.

\section{Default Data Types for Keys} \label{sec:default-valuetype}
For the cases where only one data type per key type is used, {\ViennaData} can be provided with a default data type for each key type.
This is especially common when using a type-based key dispatch for data access. Consider for example
\begin{lstlisting}
 viennadata::access<front_color, RGBColor>()(obj1) = RGBColor(255,42,0);
 viennadata::access<back_color,  RGBColor>()(obj1) = RGBColor(0,42,255);
 viennadata::access<edge_color,  RGBColor>()(obj1) = RGBColor(255,0,42);
\end{lstlisting}
Since the key classes are already named appropriately and always the some data type \lstinline|RGBColor| is used, 
setting \lstinline|RGBColor| as default data type for the respective keys allows to make the code more compact.
This can be enabled along the lines
\begin{lstlisting}
namespace viennadata
{
 namespace traits
 {
  template <>
  struct default_data_for_key<front_color>
  {
    typedef RGBColor    type;
  };
  // and similary for back_color, edge_color
 }
} 
\end{lstlisting}
Now, the initial code snippet can be written more compactly as
\begin{lstlisting}
 viennadata::access<front_color>()(obj1) = RGBColor(255,42,0);
 viennadata::access<back_color>()(obj1) = RGBColor(0,42,255);
 viennadata::access<edge_color>()(obj1) = RGBColor(255,0,42);
\end{lstlisting}
If no default data type is specified, but the second argument is nevertheless omitted, a compile time error is thrown.